\textbf{ Spectrophotometric parallaxes with linear models:\\ Accurate distances for luminous red-giant stars }

### 2018-10-14 paper/sp.tex
clean up gaia quality cuts

% To-do items
% -----------
% - Coryn's point that we should carefully explain how our model is geometric.
% - check with WISE experts that we have the correct WISE photometry
% - write words about other linear models we have made or could make, probably in the discusssion.
% - are we citing all the proper DR2 papers we need to cite?
% - HWR: Footnote to explain the use of the word ``feature'' for astronomers.
% - HWR: Figures 1 and 2 need to be combined to directly compare Parent and Training.
% - HWR: Figure 3: Note that the y-axis is positive-definite!
% - HWR: Figure 7: Caption needs to say what arrows represent.
% - Make the comparison with Cannon as a function of L1 strength and maybe look at correlation coefficients between parameter vectors.
% - Get model running on ASPCAP outputs and compare in Discussion.
% - Make sure we incorporate / respond to Ness comments from Aug 13 email.
% - Thank APOGEE telecon and add paragraph about AGB stars and the color-magnitude diagram in the discussion.
% - Describe how we ran The Cannon; discussion of how we compare to The Cannon.
% - Cite and describe Eilers et al.
% - comment on parallax offset being a function of position and color
% - remove Cannon plot for the moment
% - comment on new plot of new precision on parallax compared to Gaia's precision

@@

The model and log-likelihood function can be expressed heuristically as
\begin{eqnarray}
\gparallax_n &=& \exp(\theta\cdot x_n) + \mbox{noise}
\\
\chi^2(\theta) &\equiv& \sum_{n=1}^N \frac{[\gparallax_n - \exp(\theta\cdot x_n)]^2}{\sigma_n^2}
\quad ,
\end{eqnarray}
where
$\gparallax_n$ is the \gaia\ measurement (adjusted; see below) of the astrometric parallax of star $n$,
the model is that the logarithm of the true parallax
can be expressed as a linear combination of the components
of some $D$-dimensional feature vector $x_n$,
$\theta$ is a $D$-vector of linear coefficients,
$\sigma_n$ is the \gaia\ estimate of the uncertainty on the parallax measurement,
and $\chi^2(\theta)$ is (twice) the negative-log-likelihood for the parameters $\theta$
under the assumption of known Gaussian noise and
that there are $N$ independently measured stars $n$.
The feature vector $x_n$ contains photometry in a few bands, and all 7400-ish pixels
in the pseudo-continuum-normalized \apogee\ spectrum, so $D$ is on the order of 7400.

In addition, we assume that many entries in the parameter $D$-vector $\theta$ will be zero
(the ``sparsity'' assumption).
In order to encourage this, we optimize not $\chi^2$ but a regularized objective function
\begin{eqnarray}
\hat{\theta} &\leftarrow& \argmin_{\theta}\left[\frac{1}{2}\,\chi^2(\theta) + \lambda\,||P\cdot\theta||_1^1\right]
\label{eq:objective}\quad ,
\end{eqnarray}
where
$\lambda$ is a regularization parameter,
$P$ is a projection operator that selects out from the $\theta$ vector only those components
that pertain to the \apogee\ spectral pixels,
and $||y||_1^1$ is the L1-norm or sum of absolute values of the components of $y$.
This kind of regularization (L1) adds a convex term to the optimization and leads to
sparse optima.
The $P$ operator makes it such that we only regularize the components of $\theta$ that multiply
the spectral pixels; we ask for the spectral model to be sparse but we don't ask for the photometric
model to be sparse.
We set the value of $\lambda$ by cross-validation.

@@

Once the model is optimized, the output of the model is a prediction of the parallax,
or really what we will call the \emph{spectrophotometric parallax}.
This spectrophotometric parallax
$\sparallax_m$ for any star $m$ in the validation or test set is
assigned according to
\begin{eqnarray}
\sparallax_m &\leftarrow& \exp(\hat{\theta}\cdot x_m)
\quad ,
\end{eqnarray}
where
$\hat{\theta}$ is the optimal parameter vector according
to \equationname~(\ref{eq:objective}),
and
$x_m$ is the feature $D$-vector for star $m$.
Thus the model can be trained on a training set of stars and applied to
a validation set or a test set to make predictions.
The only requirements are that every test-set object $m$  must have a full feature
vector $x_m$ just as every training-set object $n$ must have a full feature
vector $x_n$.

@@

For training, we need A and B Training Sets.
We define the Training Set stars to be Parent-Sample stars that also
have a measured \gaia\ parallax.
The Training-Set stars, in addition to meeting all Parent-Sample cuts,
also must meet $\sigma_{\varpi^{(\rm a)}} < 0.1$, it must be observed by \gaia\ for more than $8$ individual times, i.e. \code{visibility_periods_used}~$\geq~8$, as well as meet the additional quality criterion $\code{astrometric_chi2_al}/\sqrt{\code{astrometric_n_good_obs_al}-5} \leq 35$, which eliminates detections that are not consistent with a single PSF \citep[see][]{calj}, to ensure that the parallax
measurements are good.
These cuts leave 28226 stars total in the Training Sets.
But---as we have emphasized above---we do not cut ever on parallax or
parallax error or their ratio. For this reason, most of the training set
do not have significantly measured parallaxes and more than 2~percent even have
negative parallaxes!
But we will perform our training such that it will be unbiased in these
circumstances.
Indeed it would be necessarily biased if ever we did cut on parallax or
parallax signal-to-noise.

@@

For every star $n$ in the full Parent Sample we construct the feature
vector $x_n$ as
\begin{eqnarray}
x_n\T &\equiv& [1, G_n, \BP_n, \RP_n, J_n, H_n, K_n, W_{1n}, W_{2n}, \ln f_{1n}, \ln f_{2n}, \cdots, \ln f_{Ln}]
\quad ,
\end{eqnarray}
where the 1 permits a linear offset,
the photometry is from \gaia, \zmass, and \wise, respectively,
the fluxes in the $L=7405$ \apogee\ spectral pixels (for which there are reliably
and consistently data) for star $n$ are denoted
$f_{1n}$, $f_{2n}$, and so on,
and we have taken logarithms of those fluxes.
These feature vectors live in a $D$-dimensional space where $D=7414$.
For every star $n$ in either of the Training Sets we additionally require---in
addition to these feature vectors---a \gaia-measured astrometric parallax $\gparallax_n$
and uncertainty $\sigma_n$.

>>>>

% To-do items
% -----------
% - Coryn's point that we should carefully explain how our model is geometric.
% - check with WISE experts that we have the correct WISE photometry
% - write words about other linear models we have made or could make, probably in the discusssion.
% - are we citing all the proper DR2 papers we need to cite?
% - HWR: Footnote to explain the use of the word ``feature'' for astronomers.
% - HWR: Figures 1 and 2 need to be combined to directly compare Parent and Training.
% - HWR: Figure 3: Note that the y-axis is positive-definite!
% - HWR: Figure 7: Caption needs to say what arrows represent.
% - Make the comparison with Cannon as a function of L1 strength and maybe look at correlation coefficients between parameter vectors.
% - Get model running on ASPCAP outputs and compare in Discussion.
% - Make sure we incorporate / respond to Ness comments from Aug 13 email. And thank Ness.
% - Thank APOGEE telecon and add paragraph about AGB stars and the color-magnitude diagram in the discussion.
% - Describe how we ran The Cannon; discussion of how we compare to The Cannon.
% - Cite and describe Eilers et al.
% - comment on parallax offset being a function of position and color
% - remove Cannon plot for the moment
% - comment on new plot of new precision on parallax compared to Gaia's precision
% - Get figure showing how much better we are than Gaia at large distance and as a function of distance. And write a paragraph about said figure.

@@

The model and log-likelihood function can be expressed heuristically as
\begin{eqnarray}
\gparallax_n &=& \exp(\theta\cdot x_n) + \mbox{noise}
\\
\chi^2(\theta) &\equiv& \sum_{n=1}^N \frac{[\gparallax_n - \exp(\theta\cdot x_n)]^2}{\sigma_n^2}
\end{eqnarray}
where
$\gparallax_n$ is the \gaia\ measurement (adjusted; see below) of the astrometric parallax of star $n$,
the model is that the logarithm of the true parallax
can be expressed as a linear combination of the components
of some $D$-dimensional feature vector $x_n$,
$\theta$ is a $D$-vector of linear coefficients,
$\sigma_n$ is the \gaia\ estimate of the uncertainty on the parallax measurement,
and $\chi^2(\theta)$ is (twice) the negative-log-likelihood for the parameters $\theta$
under the assumption of known Gaussian noise and
that there are $N$ independently measured stars $n$.
The feature vector $x_n$ contains photometry in a few bands, and all 7400-ish pixels
in the pseudo-continuum-normalized \apogee\ spectrum, so $D$ is on the order of 7400.

In addition, we assume that many entries in the parameter $D$-vector $\theta$ will be zero
(the ``sparsity'' assumption).
In order to encourage this, we optimize not $\chi^2$ but a regularized objective function
\begin{eqnarray}
\hat{\theta} &\leftarrow& \argmin_{\theta}\left[\frac{1}{2}\,\chi^2(\theta) + \lambda\,||P\cdot\theta||_1^1\right]
\label{eq:objective}
\end{eqnarray}
where
$\lambda$ is a regularization parameter,
$P$ is a projection operator that selects out from the $\theta$ vector only those components
that pertain to the \apogee\ spectral pixels,
and $||y||_1^1$ is the L1-norm or sum of absolute values of the components of $y$.
This kind of regularization (L1) adds a convex term to the optimization and leads to
sparse optima.
The $P$ operator makes it such that we only regularize the components of $\theta$ that multiply
the spectral pixels; we ask for the spectral model to be sparse but we don't ask for the photometric
model to be sparse.
We set the value of $\lambda$ by cross-validation.

@@

Once the model is optimized, the output of the model is a prediction of the parallax,
or really what we will call the \emph{spectrophotometric parallax}.
This spectrophotometric parallax
$\sparallax_m$ for any star $m$ in the validation or test set is
assigned according to
\begin{eqnarray}
\sparallax_m &\leftarrow& \exp(\hat{\theta}\cdot x_m)
\end{eqnarray}
where
$\hat{\theta}$ is the optimal parameter vector according
to \equationname~(\ref{eq:objective}),
and
$x_m$ is the feature $D$-vector for star $m$.
Thus the model can be trained on a training set of stars and applied to
a validation set or a test set to make predictions.
The only requirements are that every test-set object $m$  must have a full feature
vector $x_m$ just as every training-set object $n$ must have a full feature
vector $x_n$.

@@

For training, we need A and B Training Sets.
We define the Training Set stars to be Parent-Sample stars that also
have a measured \gaia\ parallax.
The Training-Set stars, in addition to meeting all Parent-Sample cuts,
also must meet $\sigma_{\varpi^{(\rm a)}} < 0.1$, it must be observed by
more than $8$ (widely separated) times
\begin{equation}
\code{visibility_periods_used >= 8}
\end{equation}
as well as meet the additional quality criterion
\begin{equation}
\code{astrometric_chi2_al / sqrt(astrometric_n_good_obs_al - 5) <= 35}
\end{equation}
which eliminates detections that are not consistent with a single PSF \citep[see][]{calj},
to ensure that the parallax measurements are good.
These cuts leave 28226 stars total in the Training Sets.
But---as we have emphasized above---we do not cut ever on parallax or
parallax error or their ratio. For this reason, most of the Training-Set stars
do not have significantly measured parallaxes and more than 2~percent even have
negative parallaxes!
But we will perform our training such that it will be unbiased in these
circumstances.
Indeed it would be necessarily biased if ever we did cut on parallax or
parallax signal-to-noise.

@@

For every star $n$ in the full Parent Sample we construct the feature
vector $x_n$ as
\begin{eqnarray}
x_n\T &\equiv& [1, G_n, \BP_n, \RP_n, J_n, H_n, K_n, W_{1n}, W_{2n}, \ln f_{1n}, \ln f_{2n}, \cdots, \ln f_{Ln}]
\end{eqnarray}
where the 1 permits a linear offset,
the photometry is from \gaia, \zmass, and \wise, respectively,
the fluxes in the $L=7405$ \apogee\ spectral pixels (for which there are reliably
and consistently data) for star $n$ are denoted
$f_{1n}$, $f_{2n}$, and so on,
and we have taken logarithms of those fluxes.
These feature vectors live in a $D$-dimensional space where $D=7414$.
For every star $n$ in either of the Training Sets we additionally require---in
addition to these feature vectors---a \gaia-measured astrometric parallax $\gparallax_n$
and uncertainty $\sigma_n$.
